Optimizing a PoS Tag Set for Dependency Parsing

### 2016-05-07 submission.tex
fixed typos, minor changes

In this paper, we optimize a PoS tag set for the task of data-driven dependency
parsing of Norwegian. We report on a set of experiments where various
morphological distinctions are introduced to the PoS annotations and evaluated
both intrinsically, i.e., in terms of PoS tagging accuracy, and extrinsically,
in terms of parsing accuracy. Our results show that the introduction of
morphological distinctions not present in the original tag set, whilst
compromising tagger accuracy, actually leads to significantly improved parsing
accuracy. This optimization also allows us to bypass the additional step of
morphological analysis, framing the whole preprocessing problem as a simple
tagging task.

@@

In the current paper, we argue that it is nevertheless possible to significantly improve the performance
of the downstream application of syntactic parsing by introducing
linguistically informed, syntactically informative distinctions to a tag set.
In the following section we first turn to the treebank that provides the basis for our experiments and
outline its morphological and syntactic annotation choices.

@@

\paragraph{Predicted vs gold tags}
As we seek to quantify the effects of PoS tagging in a realistic setting, i.e.,
in application to raw text, we choose to evaluate the parser on automatically
predicted PoS tags. For training, however, we have two options: using either
gold standard tags or automatically predicted tags. In order to settle on a
configuration, we conducted experiments with training on both gold tags 
and predicted tags to see how the performance differs. The
results of these experiments, shown in Table \ref{tagconfigeval}, reveal that
the combination of training and testing on predicted tags is clearly superior
to training on gold standard tags and testing on predicted tags. Consequently,
the parser is both trained and tested on automatically predicted tags in the
remainder of our experiments.\footnote{Though some parsers also make use of
  morphological features, we removed all morphological features beyond the PoS
  tags in order to simulate a realistic setting.}

@@

The modified tag sets used in the experiments reported in this section are
defined as combinations of PoS tags and morphological features already
available in the gold annotations of the treebank. In effect, we redefine the
tags comprising the gold standard as provided by NDT. The best performing
configuration can be seen as tag set specifically optimized for dependency
parsing. Note that the tag selection process itself can be seen as
semi-automatic; while we for each introduced distinction empirically evaluate
its impact on parsing performance -- as to see whether it is worth including in
the final configuration -- the process is manually guided by linguistic
considerations regarding which combinations to evaluate in the first place.
Although our expressed goal is to identify a tag set optimized for the
downstream task of dependency parsing, we will only consider tag set
modifications we deem linguistically sensible.

The hope is that the introduction of more fine-grained linguistically motivated
distinctions in the tag set may assist the parser in recognizing and
generalizing syntactic patterns.  However, this would also increase the
complexity of the tagging task, which can be expected to lead to a drop in
tagger accuracy.  However, the best tagging, evaluated in isolation, does not
necessarily lead to the best parse, hence the aim of this section is to
investigate how tag set modifications affect this interplay.

@@

\subsection{Tag Set Experiments}
We modify the tags for nouns (\texttt{subst}), verbs (\texttt{verb}),
adjectives (\texttt{adj}), determiners (\texttt{det}) and pronouns
(\texttt{pron}) in NDT by appending selected sets of morphological features to
each tag in order to increase the linguistic information expressed by the tags.
For each tag, we in turn first experiment with each of the available features in
isolation before testing various combinations. We base our choices
of combinations on how promising the features are and what we deem worth
For each tag, we first experiment with each of the available features in
isolation before employing various combinations of them. We base our choices of
combinations on how promising the features are and what we deem worth
investigating in terms of linguistic utility, in order to see how the features
might interact.

@@

\paragraph{Pronouns}
Pronouns in Norwegian are personal, reciprocal, reflexive and interrogative.
They can furthermore exhibit gender, number and person, while personal pronouns
can be distinguished on case (either accusative or nominative).

@@

The most successful tag set modification for each PoS and the results from
tagging and parsing with the respective modifications are seen in Table
\ref{respectiveresults}. Nouns benefit by far the most from the introduction of
more fine-grained linguistically motivated distinctions, with an LAS of 88.81\%
and UAS of 91.73\% when distinguishing on type, case and definiteness. We
observe that the most promising tag set modifications for verbs, adjectives,
determiners and pronouns all reach LAS of \textasciitilde 87.30\% and UAS of
\textasciitilde 90.40\%. To investigate the overall effect of these tag set
modifications, we tested each of the improvements in parser accuracy scores
from that of the original tag set for statistical significance using Dan
Bikel's randomized parsing evaluation comparator script\footnote{Available as
    \texttt{compare.pl} at \url{http://ilk.uvt.nl/conll/software.html}}, as
used in the CoNLL shared tasks. For the most successful tag set modification
for each of the categories seen in Table \ref{respectiveresults}, the
difference in LAS from the original tag set is statistically significant at
significance level 0.05 (\emph{p}-value < 0.05), as are all differences in UAS,
except for verbs with finiteness (\emph{p}-value 0.15) and pronouns with type
and case (\emph{p}-value 0.06).

>>>>

In this paper, we optimize a PoS tag set for the task of data-driven dependency
parsing of Norwegian. We report on a set of experiments where various
morphological distinctions are introduced to the PoS annotations and evaluated
both intrinsically, i.e., in terms of PoS tagging accuracy, and extrinsically,
in terms of syntactic parsing accuracy. Our results show that the introduction
of morphological distinctions not present in the original tag set, whilst
compromising tagger accuracy, actually leads to significantly improved parsing
accuracy. This optimization also allows us to bypass the additional step of
morphological analysis, framing the whole preprocessing problem as a simple
tagging task.

@@

In the current paper, we argue that it is nevertheless possible to
significantly improve the performance of the downstream application of
syntactic parsing by introducing linguistically informed, syntactically
informative distinctions to a tag set. In the following section, we first turn
to the treebank that provides the basis for our experiments and outline its
morphological and syntactic annotation choices.

@@

\paragraph{Predicted vs. Gold Tags}
As we seek to quantify the effects of PoS tagging in a realistic setting, i.e.,
in application to raw text, we choose to evaluate the parser on automatically
predicted PoS tags. For training, however, we have two options: using either
gold standard tags or automatically predicted tags. In order to settle on a
configuration, we conducted experiments with training on both gold tags
and predicted tags to see how the performance differs. The
results of these experiments, shown in Table \ref{tagconfigeval}, reveal that
the combination of training and testing on predicted tags is clearly superior
to training on gold standard tags and testing on predicted tags. Consequently,
the parser is both trained and tested on automatically predicted tags in the
remainder of our experiments.\footnote{Though some parsers also make use of
  morphological features, we removed all morphological features beyond the PoS
  tags in order to simulate a realistic setting.}

@@

The modified tag sets used in the experiments reported in this section are
defined as combinations of PoS tags and morphological features already
available in the gold annotations of the treebank. In effect, we redefine the
tags comprising the gold standard as provided by NDT. The best performing
configuration can be seen as a tag set specifically optimized for dependency
parsing. Note that the tag selection process itself can be seen as
semi-automatic; while we for each introduced distinction empirically evaluate
its impact on parsing performance -- as to see whether it is worth including in
the final configuration -- the process is manually guided by linguistic
considerations regarding which combinations to evaluate in the first place.
Although our expressed goal is to identify a tag set optimized for the
downstream task of dependency parsing, we will only consider tag set
modifications we deem linguistically sensible.

The hope is that the introduction of more fine-grained linguistically motivated
distinctions in the tag set may assist the parser in recognizing and
generalizing syntactic patterns.  However, this would also increase the
complexity of the tagging task, which can be expected to lead to a drop in
tagger accuracy.  The best tagging, evaluated in isolation, does not
necessarily lead to the best parse, hence the aim of this section is to
investigate how tag set modifications affect this interplay.

@@

\subsection{Tag Set Experiments}
We modify the tags for nouns (\texttt{subst}), verbs (\texttt{verb}),
adjectives (\texttt{adj}), determiners (\texttt{det}) and pronouns
(\texttt{pron}) in NDT by appending selected sets of morphological features to
each tag in order to increase the linguistic information expressed by the tags.
For each tag, we in turn first experiment with each of the available features in
isolation before testing various combinations.
%We base our choices
%of combinations on how promising the features are and what we deem worth
We base our choices of
combinations on how promising the features are and what we deem worth
investigating in terms of linguistic utility, in order to see how the features
might interact.

@@

\paragraph{Pronouns}
Pronouns in Norwegian include personal, reciprocal, reflexive and
interrogative.  They can furthermore exhibit gender, number and person, while
personal pronouns can be distinguished on case (either accusative or
nominative).

@@

The most successful tag set modification for each PoS and the results from
tagging and parsing with the respective modifications are seen in Table
\ref{respectiveresults}. Nouns benefit by far the most from the introduction of
more fine-grained linguistically motivated distinctions, with an LAS of 88.81\%
and UAS of 91.73\% when distinguishing on type, case and definiteness. We
observe that the most promising tag set modifications for verbs, adjectives,
determiners and pronouns all reach LAS of \textasciitilde 87.30\% and UAS of
\textasciitilde 90.40\%. To investigate the overall effect of these tag set
modifications, we tested each of the improvements in parser accuracy scores
from those of the original tag set for statistical significance using Dan
Bikel's randomized parsing evaluation comparator script\footnote{Available as
    \texttt{compare.pl} at \url{http://ilk.uvt.nl/conll/software.html}}, as
used in the CoNLL shared tasks. For the most successful tag set modification
for each of the categories seen in Table \ref{respectiveresults}, the
difference in LAS from the original tag set is statistically significant at
significance level 0.05 (\emph{p}-value < 0.05), as are all differences in UAS,
except for verbs with finiteness (\emph{p}-value 0.15) and pronouns with type
and case (\emph{p}-value 0.06).
